R for Clinical Study Reports and Submission

### 2023-07-16 tlf-disposition.qmd
Migrate from bookdown to Quarto book (working HTML book)
update tlf-disposition.qmd

# Disposition {#disposition}

@@

The disposition of participants table reports the numbers of participants who were randomized,
and who entered and completed each phase of the study.
In addition, the reasons for all post-randomization discontinuations,
grouped by treatment and by major reason (lost to follow-up, adverse event, poor compliance, etc.) are reported.

```{r, include=FALSE}
source("common.R")
```

@@

In this chapter, we show how to create a typical disposition table.

@@

The first step is to read in the relevant datasets into R.
For a disposition table, all the required information is saved in a Subject-level Analysis Dataset (ADSL).
This dataset is provided in `sas7bdat` format, which is a SAS data format currently used in many clinical trial analysis and reporting.
The `haven` package is able to read the dataset, while maintaining its attributes (e.g., variable labels).

@@

The following variables are used in the preparation of a simplified disposition of participants table:

- USUBJID: unique subject identifier
- TRT01P: planned treatment
- TRT01PN: planned treatment numeric encoding
- DISCONFL: discontinued from study flag
- DCREASCD: discontinued from study reason coded

@@

In the code below, we calculate the number of participants in the analysis population by treatment arms.

@@

@@

In the code below, we calculate the number and percentage of
participants who completed/discontinued the study for different reasons by treatment arms.

@@

In the code below, we calculate the number and percentage of participants who complete the study by treatment arms.
We split `n_reason` because we want to customize the row order of the table.

@@

In the code below, we calculate the numbers and percentages of participants who discontinued the study for different reasons by treatment arms.
For display purpose, `paste0("    ", row)` is used to add leading spaces to produce indentation in the final report.

@@

Now we combine individual rows into one table for reporting purpose.
`tbl_disp` is used as input for r2rtf to create final report.

@@

In the below code, formatting of the final table is defined.
Items that were not discussed in the previous sections, are highlighted below.

The `rtf_title` defines table title.
We can provide a vector for the title argument.
Each value is a separate line.
The format can also be controlled by providing a vector input in text format.

@@

The procedure to generate a disposition table can be summarized as follows:

- Step 1: Read subject level data (i.e., `adsl`) into R.
- Step 2: Count participants in the analysis population and name the dataset `n_rand`.
- Step 3: Calculate the number and percentage of participants who discontinued the study by treatment arm, and name the dataset `n_disc`.
- Step 4: Calculate the numbers and percentages of participants who discontinued the study for different reasons by treatment arm,
  and name the dataset `n_reason`.
- Step 5: Calculate the number and percentage of participants who completed the study by treatment arm, and name the dataset `n_complete`.
- Step 6: Bind `n_rand`, `n_disc`, `n_reason`, and `n_complete` by row.
- Step 7: Write the final table to RTF

>>>>

# Disposition {#sec-disposition}

```{r, include=FALSE}
source("_common.R")
```

@@

The participant disposition table provides an overview of the study's participant numbers, 
detailing those who were randomized and their progression through each phase. 
It also presents the reasons for discontinuation after randomization, 
categorized by treatment and major factors such as 
lost to follow-up, adverse events, poor compliance, and others.

@@

This chapter provides a step-by-step guide on creating a simple disposition table in RTF format.
The expected output of the table is as follows:

@@

## Input data

The initial step involves reading the necessary datasets into R. 
In the case of a disposition table, 
the required information is typically stored in a Subject-level Analysis Dataset (ADSL). 
The dataset provided is in `sas7bdat` format, commonly used in clinical trial analysis and reporting. 
To read the dataset while preserving its attributes, such as variable labels, 
we can use the `read_sas` function in haven package.

@@

To prepare a simplified disposition of participants table using the `adsl` dataset, the following variables are utilized:

- `USUBJID`: unique subject identifier
- `TRT01P`: planned treatment
- `TRT01PN`: planned treatment numeric encoding
- `DISCONFL`: discontinued from study flag
- `DCREASCD`: discontinued from study reason

@@

## Output data

The objective of this section is to create a dataset suitable for generating the expected report. 
This process will consist of several steps to generate each component of the reporting dataset.

we can calculate the number of participants 
in the analysis population for each treatment arm by using following code.

@@

Next, we calculate the number of participants who have been discontinued from the study.

@@

Additionally, we calculate the number and percentage of participants
who completed or discontinued the study, categorized by treatment arms and different reasons.

@@

In the following code, we compute the number and percentage of participants 
who successfully complete the study for each treatment arm. 
The splitting of `n_reason` is done to allow for customizing the row order within the resulting table.

@@

The following code calculates the numbers and percentages of participants 
who discontinued the study for various reasons, categorized by treatment arms. 
To achieve indentation in the final report, 
the code `paste0("    ", row)` is employed to add leading spaces.

@@

To summarize the data manipulation steps, 
we consolidate individual rows into a single table for reporting purposes. 
This table, named `tbl_disp`, 
serves as the input for generating the final report using the `r2rtf` package.

@@

## Reporting

By using `tbl_disp` as input, we create generate the disposition table using following code.
The `r2rtf` elements that were not covered in the previous sections are highlighted.
Here, the `rtf_title` function is utilized to define the title of the table. 
For the `title` argument, 
you can provide a vector where each value represents a separate line.

@@
